---
title: "Predi-Pitch: Using Random Forests to Predict Pitch Types"
author: "David Scolari, Harrison Snell, Brandon Williams"
date: "5/09/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(here)
path <- here()

source(file.path(path, 'code', 'dugout.R'))

load(file = file.path(path, 'output', 'tables', "overall_performance_all.RDs"))


foreach(id = 1:30) %do% {
  # make filenames according to loop index
  fname_type <- paste0("types", id, ".RDs")
  fname_zones <- paste0("zones", id, ".RDs")
  fname_bypitch <- paste0("by_pitch_performance", id, ".RDs")

  # load table outputs according to loop index
  load(file = file.path(path, 'output', 'tables', fname_type))
  load(file = file.path(path, 'output', 'tables', fname_zones))
  load(file = file.path(path, 'output', 'tables', fname_bypitch))
}

knit_print.data.frame <- lemon_print
load(file.path(path, "output", "pitches_import.RData"))

```

## Introduction
Hitting a baseball has been shown to be one of the most difficult tasks in all of sports. Anybody who has gone to a batting cage and tried the fastest machine for fun knows how hard it can be to hit a normal fastball. That does not even include the possibilities of breaking balls and off-speed pitches. Batters have a fraction of a fraction of a second to recognize a pitch, determine if that pitch will be a ball or a strike, and set their swing into motion. Considering the human body can only move so fast to get the bat over the plate, the batter has to make their decision moments after the ball leaves the pitcher's hand. To give themselves the best chance of making good contact, professional baseball players will often attempt to predict whether a changeup, fastball, or a curveball is coming before the ball leaves the pitcher's hand.

Hitters will use a variety of information to inform their prediction of an upcoming pitch, including scouting reports that teams build coming into a game, situational context, memory of the previous few pitches, and game specific tendencies of a pitcher. In this project, we attempt to model this decision making process with a random forest pitch type classifier. We use pitch level data from the 2015-2018 seasons to predict pitch types for 30 different pitchers who played during that type period. 

Our model aims to contribute to the baseball world in the following ways. First, by testing how different sets of features perform at classifying out of sample pitch types, we see that some features are more important for certain pitchers than they are for others. This will help teams focus on the important information when building scouting reports going into a game against a certain pitcher. 

Second, we aim to make the results of our pitch type predictor accessible to baseball fans in hopes that it can augment their viewing experience. For many enthusiasts, strategizing along side the players from pitch to pitch, guessing both what the pitcher is going to throw as well as what the batter's strategy will be, is the most enjoyable part of the game. However, among more casual fans, baseball has the reputation for being "boring" to watch, largely due to the pauses between every pitch, during which players sometimes take 30 or more seconds to relay signs and perform ritualistic jersey adjustments, bat taps, and other manner of baseball superstitions. To make this aspect of baseball more palatable to fans, we built an R Shiny web application that allows them to input game situations and call our model's pitch type predictions in real time, allowing them to take part in the "game within the game" of baseball. 

This report proceeds as follows. First we analyze the physical attributes of the pitches that our 30 selected pitchers throw, making the case that it is indeed in the batters best interest to have an idea of how a particular pitch is going to move once it leave the pitcher's hand. Second, we show some evidence that pitchers do adopt different strategies for different game situations. We then describe the rich, pitch level data set we use for this analysis. Finally, we discuss our random forest classifier and the results it produces. 

### Reproducibility

While much of our analysis can be reproduced directly from the RMarkdown file, there are several scripts in our Github that are called at various times. In order to have 100% reproducibility, the following load order can be used:

The following scripts allow the writeup to be reproduced:

* dugout.R (similar to a include.R) loads the necessary libraries and establishes the file path
*  **import.R** converts the enormous dataset into a more manageable subset of MLB pitchers
*  **pitchers.R** pre-processes the data so that it is easily fed into the predipitch script
*  **predi_pitch.R** takes the subset of pitchers and creates a predictive model for each one (**NOTE:** Be advised, this will take upwards of 30 minutes to run)
* performance.R creates a table that shows the performance of the models
* kershaw_sequence.R looks at one pitcher and shows sequencing trends throughout the game in different scenarios
* situation.R allows the manual entry of at-bat scenarios (balls, strikes, etc.) to create a prediction
* predipitch.Rmd creates the final write-up with visualizations

Due to the size of the datasets at hand, running the scripts in **bold** can take a long time. However, they output .RDs which can referenced directly from the Github so you may skip these scripts and run only the other scripts of interest. Just note, dugout.R should be run in either case. 

Finally, due to the size of datasets, we gitignore all .csv files. You can find some of the data in archive.zip. However, download pitches.zip from one of the links in the appendix to complete your local repository if you wish to use the import.R script.

## The Data
The data involved in this project consists of every pitch from the 2015 through the 2018 Major League Baseball (MLB) seasons. These four seasons contained nearly 3 million observations, and the data include categorization of 16 different types of pitches, the spin, speed, and location of each pitch, the game situation (score, runners on base, balls and strikes, etc.), information about the pitcher and batter, and the result of the at-bat. 

From this data set, we engineer a few important features. Since each pitch is identified by an at bat id and a pitch number, we can add the previous pitch types as features to each pitch. For this model, we add categorical features for the previous two pitches. We also add the previous at bat's result as a feature.

We also add a factor variable indicating the position in the batting order the pitcher is facing. A team's best hitters tend to bat 3rd or 4th in the order, and a team's worst hitter tends to bat 8th and 9th, so batting order may be able to tell our model a lot about what pitch types a pitcher will throw. 

Lastly, we add features for the game specific share of a pitcher's total pitches that each pitch types makes up. This set of features is meant to capture the fact that a pitcher's arsenal of pitch types may change from game to game. Most games a pitcher might throw mostly four seam fastballs, but there might be game specific "fixed effects" that lead him to rely more on his secondary pitches. We'd like our model to be able to use this information in making its predictions.

## Why Do We Care About What Pitch is Coming?
### The Difference Between Pitch Types
Not all pitches are made the same. From sliders to cutters to knuckleballs, each pitch tends to have its own characteristics that define it (see Appendix E for a list of pitch type acronyms). From the following table, we see just how vastly different each pitch can be. On average we find that fastballs start at almost 94 mph and changeups start at 86 mph, a massive difference considering the pitches are designed to look the exact same to the batter. The average break length of a fastball is about 4 inches, which means the point that a batter sees a fastball leave the pitcher's hand will follow a mostly straight line. Compare that to a curveball with almost 13 inches of break and we start to see why knowing the next pitch can be helpful.

```{r caption = "Averages of Different Pitch Characteristics", render=lemon_print}

pitch_sum = pitches %>%
  filter(pitch_type != "") %>%
  filter(pitch_type != "IN")%>%
  dplyr::select(pitch_type,px,pz,start_speed,end_speed,spin_rate,spin_dir,break_angle,break_length)%>%
  mutate(pitch_type = as.factor(pitch_type))%>%
  group_by(pitch_type)%>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))

pitch_sum = pitch_sum %>%
  mutate_if(is.numeric,round, digits=3)

pitch_sum
```

The following plots of pitcher Chris Archer's pitches allow us to visualize the differences in each pitch type. For Archer, we see that his slider tends to have a relatively low spin rate, with his fastballs having higher spin rates and his changeups in between. As we would expect, Archer's fastballs have little break. His slider, however, has a large but variable amount of break and always tends to break in the same direction. These visuals help us understand where a pitch may be going. If we can predict the next pitch is a slider, then we can start to understand the direction and magnitude of break in the pitch. 



```{r fig.height = 4}

plot_data = pitches %>%
  filter(pitch_type != "") %>%
  filter(pitch_type != "IN")%>%
  filter(pitch_type != "UN")%>%
  filter(pitch_type != "FA")%>%
  dplyr::select(pitcher_id,pitch_type,px,pz,start_speed,end_speed,spin_rate,spin_dir,break_angle,break_length,break_y)

plot_data$pitch_type = as.factor(plot_data$pitch_type)

plot_archer = plot_data%>%
  filter(pitcher_id == 502042)


ggplot(plot_archer)+
  geom_point(aes(x=spin_rate,y=spin_dir,color=pitch_type),alpha=.2)+
  ggtitle("Spin Rate and Direction for Chris Archer's Pitches")
```

```{r fig.height = 4}
ggplot(plot_archer)+
  geom_point(aes(x=break_angle,y=break_length,color=pitch_type),alpha=.2)+
  ggtitle("Break Angle and Length for Chris Archer's Pitches")
```


```{r fig.height = 4}
ggplot(plot_data)+
  geom_density(aes(x=start_speed), fill="blue")+
  ggtitle("Distribution of Start Speeds across Pitches")+
  facet_wrap(~pitch_type)
```

```{r fig.height = 4}

ggplot(plot_data)+
  geom_density(aes(x=break_length), fill="red")+
  ggtitle("Distribution of Break Length across Pitches")+
  facet_wrap(~pitch_type)
```

Mostly as an exploration into our data, we found the graphs of distributions of break length and start speed to be quite interesting. They provide a  thorough comparison of pitch types because every pitcher is slightly different. Some pitchers may have a lot more movement on their sinker than others. Similarly, Aroldis Chapman's legendary 105 mph fastballs are certainly not the norm in the league.  In Appendix B, we present the pitching profiles of the 30 pitchers based on the style of their pitches and the frequency that they are thrown. Getting to see the variability of break length and speeds begins to paint the picture of why some pitchers are better than others, and why predicting what is coming is so important. 


### The Difference Between Situations
Naturally, a pitcher will have a generally well established "arsenal" of pitch types that he throws on a regular basis, so a batter will have an idea of types of pitches he is likely to face and the relative frequency at which he will face them. However, by being aware of the game situation, a batter might have a better idea of the type of pitch he is likely to see next. The following section illustrates how this works in practice. 

The below table lists the distribution of pitch types that Clayton Kershaw throws conditional on a few different game situations. We observe that Kershaw throws about 48% four seam fastballs across all at bats in our data set. However, if we only look at the first pitch, we see that he throws about 68% four seam fastballs. Following fastballs, his frequency of sliders and curves increases significantly. In two strikes situations, he tends to throw a lot of sliders and relied more on his curve relative to other counts. With three balls, Kershaw relies on four seam fastballs and sliders (probably his two most reliable pitches). To the "heart" of the order, or the three and four hitters in the batting order, he ups the frequency of sliders and curves and throws less four seamers. To the bottom of the order (eight and nine hitters), he throws more four seamers. He also throws a lot of fastballs to lefties, probably because lefties hit off left handed pitchers with a lower batting average than righties do. 

```{r}

load(file = file.path(path, 'output', 'tables', 'kershaw_sequence.RData'))
pitch_sequence

```

We can further improve the specificity of our distribution by interacting elements of a situation. For example, the below table looks at what Kershaw throws with his first pitch after a walk to the clean-up hitter (fourth in the order) and the eight and ninth place hitters. We see that after a walk, if Kershaw is facing the bottom of the order, he does not mess around very much with off speed pitches and relies heavily on the four seam. However, to clean-up hitters, Kershaw still mixes in a large share of sliders, clearly balancing the need to throw strikes after a walk with respect for the offensive threat of the clean-up hitter. These kinds of interactions are exactly what our random forest aims to leverage to make it's predictions. 

```{r}

just_walked

```


## Methodology

Given that we've established how difficult it is to get hits consistently at the major league level, having an idea an idea of what the opposing pitcher is about to throw would confer a significant advantage to the hitter. We endeavor to create a model that successfully predicts the next pitch a pitcher will throw in an at-bat, given the circumstances of the at-bat, the tendencies of the pitcher, and the progress of the game up to that point. Using three distinct random forest models, we derive a predictive approach that generally outperforms guessing that the pitcher will throw their most common pitch (often called "sitting on a pitch"), and in most cases, significantly exceeds this "sitting on a pitch" approach.

The models are generated with a train/test split and then evaluated for the out-of-sample performance against the testing set. Models are generated per pitcher, in such a way that be beneficial in application to a baseball manager or hitter, given the situation in the game. 

### Our Random Forests

Random forests, much like actual forests, are an aggregation of individual trees. Tree models make predictions through a series of binary decisions based on a selection of features that sort the data into groups of most likely outcomes. While trees are intuitive, a well performing model will add some layers of complexity that reduce over fitting. Bootstrap aggregating, or "bagging" involves taking $B$ bootstrapped samples of the original data and fitting a tree model to each one. Predictions are generated using a summary of the $B$ tree models. For a categorical outcome, each tree model contributes one vote and the outcome is classified by majority rule. 

Random forests extend bagged trees by only allowing each individual tree to use $m$ of the total number of features, $p$. So bagging is equivalent to a random forest with $m = p$. Restricting the each tree to a subset of the total features decorrelates the individual tree models, reducing variance in the predictions and making it less likely that the model will over fit the data. For our random forest models, we use a common choice of feature size, $m \approx \sqrt p$.

#### The Situational Model

The first of our three random forest models uses information readily available in the at-bat to predict the upcoming pitch. The features involved here include the ball-strike count, opposing batter's stance, inning, how many pitches thrown in the at-bat so far, the game score, and the runners on base. This is the most interpretable model, as it is composed of the factors that are generally considered most relevant and well-known by players and coaches in the moment. As random forests are an aggregation of individual trees, it can be illustrative to look at a single tree to get a sense for how decisions are being made at various nodes. Consider, for example, the following tree as an example of what the random forest is doing for pitcher Felix Hernandez. 

```{r}

## Load in data
set.seed(117) #set seed for image replication 

load(file.path(path, "output", "pitches_import.RData"))

## Felix Hernandez Example ID 2

id = 2

fname_pitcher <- paste0("pitcher", id, ".Rds")
load(file = file.path(path, 'output', 'pitchers', fname_pitcher))

fname_factor <- paste0("pitcher_factor", id, ".RDs")
load(file = file.path(path, 'output', 'pitchers', fname_factor))

pitcher_first <- names$first_name[id]
pitcher_last <- names$last_name[id]

## Train / Test

pitcher_split =  initial_split(pitcher, prop=0.8)
pitcher_train = training(pitcher_split)
pitcher_test  = testing(pitcher_split)

## Single tree and 1SE Prune

pitcher.tree = rpart(pitch_type ~ 
                       inning + 
                       b_count + 
                       s_count +
                       pitch_num +
                       stand +
                       on_1b + 
                       on_2b + 
                       on_3b +
                       outs +
                       p_score +
                       b_score,
                     data=pitcher_train,
                     control = rpart.control(cp = .005, minsplit = 30))

prune_1se = function(my_tree) {
  out = as.data.frame(my_tree$cptable)
  thresh = min(out$xerror + out$xstd)
  cp_opt = max(out$CP[out$xerror <= thresh])
  prune(my_tree, cp=cp_opt)
}

pitcher.tree_prune = prune_1se(pitcher.tree)

## Plot

rpart.plot(pitcher.tree_prune, type=4, extra=1)


```

Hernandez is a pitcher with a diverse arsenal of pitches. As one can see, the predictive model analyzes factors about the game situation to indicate the next pitch. It considers, for example, the count on the batter and whether the batter stands left or right. After running through the branches and nodes with binary decisions based on the in-game situation, a prediction about the next pitch is made at the bottom of the tree. The random forest takes a bootstrap aggregate of trees like this one, but a single tree helps us see an example of the predictive process. 

#### The Lagged Model


The second random forest builds upon the features selected in the Situational Model and supplements them with information about the previous two pitches and the event the last at-bat. This allows for the model to incorporate lagged information that might directly influence the next pitch. Did the pitcher just give up a home run on the curve? Maybe it's a steady diet of fastballs from here on out. Did the pitcher just give the batter two straight fastballs to study? Perhaps it's time for an off-speed pitch like a change-up. 

#### Trash Can Model


The third and final random forest uses nearly every feature from the dataset to control for all possible scenarios and variations in the feature matrix (with our sincerest apologies to Astros fans for the name). This means that it not only considers all the factors included in the other two models, but also includes what the pitcher's game has looked like so far; that is, it takes into account the pitch choices as a percent of the overall pitches in the game up to that point. Therefore, if the pitcher is leaning heavily on the slider that day, for example, this model will incorporate that pattern. 

### Pitch Prediction

Now that we have our models constructed, how does it work? Let's return to Clayton Kershaw in a real game situation to find out. Let's say the count is one ball, one strike, and there is one out. Kershaw is facing a right-handed batter. What does the Trash Can model say that Kershaw will throw here? 

``` {r}

# Enter the game situation 

id = 12 #Kershaw
stance = "R"
strikes = 1
balls = 1
out = 1

# Call the pitcher's dataset 

load(file.path(path, "output", "pitches_import.RData"))

fname_model <- paste0("trashcans", id, ".RData")
load(file = file.path(path, 'output', 'models', fname_model))

fname_pitcher <- paste0("pitcher", id, ".Rds")
load(file = file.path(path, 'output', 'pitchers', fname_pitcher))

fname_factor <- paste0("pitcher_factor", id, ".RDs")
load(file = file.path(path, 'output', 'pitchers', fname_factor))

pitcher_first <- names$first_name[id]
pitcher_last <- names$last_name[id]

# affix the model prediction results to the pitches dataset

pitcher$pitch_type <- factor(pitcher$pitch_type, factor_types)
pitcher_predi <- pitcher %>% 
  mutate(pitchhat_trashcan = predict(trashcan_3, pitcher))
pitcher_predi$pitchhat_trashcan = factor(pitcher_predi$pitchhat_trashcan, factor_types)

# filter by game situation

  pitcher_result <- pitcher_predi %>% 
    filter(
      stand == stance &          
        s_count == strikes &
           b_count == balls &
           outs == out
           )

# summarize the predictions and the actual results 
  
  pitcher_guess <- pitcher_result %>% 
    group_by(pitchhat_trashcan) %>% 
    summarize(prediction = n()) %>% 
    mutate(percent = prediction / sum(prediction)) %>% 
    rename(pitch_type = pitchhat_trashcan)
  
  pitcher_real <-  pitcher_result %>% 
    group_by(pitch_type) %>% 
    summarize(actual = n()) %>% 
    mutate(actual_percent = actual / sum(actual))

  situation <- merge(pitcher_guess, pitcher_real)

  situation <-situation %>% 
    arrange(desc(percent)) %>% 
    kable(caption = "Kershaw with 1 Ball, 1 Strike, 1 Out")
  
situation 

```

The model says that there's about a third chance that he will throw a slider, a third chance he will throw a fastball, and a third chance he will throw a curveball. Obviously, this is a difficult predicament for a hitter, but the model itself did pretty well! We slightly overpredict the chance of a slider and a fastball, and slightly underpredict the curveball, but our distribution is largely the same. 

What if we change the situation again? Now say it's the first pitch of an at-bat with two outs against a lefty. 

``` {r}

# Enter the game situation 

id = 12
stance = "L"
strikes = 0
balls = 0
out = 2

# Call the pitcher's dataset 

load(file.path(path, "output", "pitches_import.RData"))

fname_model <- paste0("trashcans", id, ".RData")
load(file = file.path(path, 'output', 'models', fname_model))

fname_pitcher <- paste0("pitcher", id, ".Rds")
load(file = file.path(path, 'output', 'pitchers', fname_pitcher))

fname_factor <- paste0("pitcher_factor", id, ".RDs")
load(file = file.path(path, 'output', 'pitchers', fname_factor))

pitcher_first <- names$first_name[id]
pitcher_last <- names$last_name[id]

# affix the model prediction results to the pitches dataset

pitcher$pitch_type <- factor(pitcher$pitch_type, factor_types)
pitcher_predi <- pitcher %>% 
  mutate(pitchhat_trashcan = predict(trashcan_3, pitcher))
pitcher_predi$pitchhat_trashcan = factor(pitcher_predi$pitchhat_trashcan, factor_types)

# filter by game situation

  pitcher_result <- pitcher_predi %>% 
    filter(
      stand == stance &          
        s_count == strikes &
           b_count == balls &
           outs == out
           )

# summarize the predictions and the actual results 
  
  pitcher_guess <- pitcher_result %>% 
    group_by(pitchhat_trashcan) %>% 
    summarize(prediction = n()) %>% 
    mutate(percent = prediction / sum(prediction)) %>% 
    rename(pitch_type = pitchhat_trashcan)
  
  pitcher_real <-  pitcher_result %>% 
    group_by(pitch_type) %>% 
    summarize(actual = n()) %>% 
    mutate(actual_percent = actual / sum(actual))

  situation <- merge(pitcher_guess, pitcher_real)

  situation <-situation %>% 
    arrange(desc(percent)) %>% 
    kable(caption = "Kershaw with Two Outs Against a Left-handed Batter")
  
situation  

```

Recall that we saw that Kershaw loves to throw a fastball on the first pitch, and indeed our model predicts that (at a slightly higher rate than is actually observed.) So if you are a left-handed hitter coming into the first pitch of an at-bat, it's probably a good idea to look for a fastball. 

Notice, however, that just because our distribution generally matches the true distribution of the pitches, we aren't getting everything exactly right. The following confusion matrix shows that we are failing to predict the true pitch given a real in-game situation. While we get quite a bit right, one can see where we swing and miss on the off-diagonal. 

```{r}

# Enter the specific pitcher

id = 12

# Call the pitcher's dataset 

load(file.path(path, "output", "pitches_import.RData"))

fname_model <- paste0("trashcans", id, ".RData")
load(file = file.path(path, 'output', 'models', fname_model))

fname_pitcher <- paste0("pitcher", id, ".Rds")
load(file = file.path(path, 'output', 'pitchers', fname_pitcher))

fname_factor <- paste0("pitcher_factor", id, ".RDs")
load(file = file.path(path, 'output', 'pitchers', fname_factor))

pitcher_first <- names$first_name[id]
pitcher_last <- names$last_name[id]

# affix the model prediction results to the pitches dataset

pitcher$pitch_type <- factor(pitcher$pitch_type, factor_types)
pitcher_predi <- pitcher %>% 
  mutate(pitchhat_trashcan = predict(trashcan_3, pitcher))
pitcher_predi$pitchhat_trashcan = factor(pitcher_predi$pitchhat_trashcan, factor_types)


# Make a table of classification errors
confusion <- xtabs(~pitch_type + pitchhat_trashcan, data=pitcher_predi) 
confusion 
  


```

These predictions can be generated for any pitcher, and we invite you to explore in-game situations with our pitchers in the Shiny app found here: https://hsnell-6.shinyapps.io/DataMiningProject_PitchPrediction/ 

## Results

In Table 6 (and Appendix B), we publish our models' results for the 30 pitchers we chose to look at. We break down the performance of these models by pitch type with the tables in Appendix C. These pitchers are generally considered some of the best in the business, but we also include some pitchers from outside the top tier for completeness. Although we chose 30 pitchers to model for this project, we designed our data processing and modeling to be easily adaptable to any pitcher who threw between 2015 and 2018.

```{r}

overall_performance_all

```

To measure our models' effectiveness, we compare them to the most-common pitch thrown by the pitcher (called "sitONE") and to each other. For most pitchers, their most frequent pitch is a four seam fastball, so sitONE is sitting on fastball. The out-of-sample performance of each model is presented in for the three models. 

Most models improved in their accuracy of classifying the upcoming pitch when compared to the "sitting on the pitch" as seen in Table 4. Consider the case of Chris Sale. Sale is a perennial All-Star, and finished top-5 in Cy Young (MLB's most valuable pitcher award) votes each year of our data. If a batter were to look for his most common pitch (the two-seam fastball), he would would only be right about one third of the time. However, Sale becomes increasingly more predictable as the model incorporates more features. The Situational Model and the Lagged Model predict his next pitch at 37% and 38% respectively. Still, the Trash Can predicts his next pitch with an out-of-sample accuracy of nearly 46%, a jump of about 13%. There isn't a hitter in baseball who wouldn't want to know Sale's next pitch with a 13% increase in accuracy!

On the other hand, some pitcher profiles grew in accuracy only up through the first or second model, and then decrease as more features are added. For example, Mark Melancon and Huston Street have out-of-sample prediction accuracies that peak in the Lagged Model but fall in the Trash Can when we add more features. We present this as some evidence of over-fitting for certain pitchers. 

Not all pitchers are particularly predictable, however. Corey Kluber, a two-time Cy Young winner (one time in our data window), has such a varied arsenal of pitches that even the most predictive model (.343 out of sample accuracy) barely confers any information about the next pitch, even if it beats his sit-one rate. Further, some pitchers remained elusive for all three of the models. Trevor Rosenthal, for instance, throws a fastball on 75% of his pitches, and all three models performed worse than this "sitting on the pitch" rate.

One fairly common pattern that we notice throughout the models is as we feed more features to our random forest, the models trade accuracy in the pitchers' most frequently thrown pitch for accuracy in the secondary pitches. Take Yu Darvish as an example. We see in table Table 7 that the model predicts his pitches with greater overall accuracy with each addition of features. However, if we look at Table 8, we see that from model 2 to model 3, four seam fastball accuracy actually decreases while several of Darvish's secondary pitches get classified with much greater accuracy, particularly two seam fastballs. Specific to Darvish's case, this might be due to his "streaky" use of the two seam fastball, relying on it heavily in some games and not at all in other. Or perhaps he added this pitch to his arsenal in the middle of our data set's span, which only the game specific pitch distributions will pick up. More generally, this primary/secondary pitch accuracy trade off is a common trade off that our models make to achieve better overall performance for many of the pitchers.

```{r}
by_pitch_performance_Yu_Darvish
```

However, whether this trade off is a good one depends on a batter's hitting approach. A power hitter might aim to never miss a fastball, opting to get fooled more often on off speed pitches in exchange for the ability to crush any fastball that the pitcher throws. This hitter might prefer a modeling approach that maximizes fastball performance and that does not make the primary/secondary pitch accuracy tradeoff. 

Finally, we note that predictive power does not necessarily equate to hits. Enter Zack Britton, who led the league in saves during the 2016 season. Britton is nearly a one-pitch pitcher, throwing sinkers on 89.7% of his pitches. The Trash Can model improves this prediction a few fractions of a percent, but in either case, it's fair to say that most hitters know exactly what Britton is about to throw. Nevertheless, Britton put up an otherworldly 0.54 earned run average in 2016! That is to say, even if batters knew with near perfect clairvoyance a that sinker was on its way, making contact that leads to a hit is a whole other matter.

## Conclusion

Using a random forest classifier, we predict pitch types of 30 different pitchers using pitch level data from the 2015-2018 MLB seasons. Over our three models, we progressively roll out sets of features that reflect game situations, previous pitches and results, and game specific pitch distributions. Among the 30 pitchers, some are modeled with greater accuracy with each addition of features. Others improve with the addition of "situation" and "lagged" features but seem to be over fit by the game specific pitch distribution features. For others, all of our random forests are out performed by a model that only guesses their most frequently thrown pitch.

A theme we find in our results is that the ability of our models to predict these pitchers' pitch types does not necessarily indicate that hitters will have success in facing them. Zack Britton illustrates this well, as he is highly predictable, yet still performs well in the MLB. 

Our models provide a framework for MLB clubs to build scouting reports on opposing pitchers. By filtering our predictions by situation, teams can obtain predicted pitch distributions for pivotal game situations as they see fit. Additionally, by analyzing how the different sets of features perform, a team's hitters can get a sense of which information, situational, pitch and result history, or game to game fluctuations on a pitcher's arsenal, are relevant pieces of information to consider in their in-game mental calculus of prediction pitches. 

Beyond tools for professional teams, our models are designed to be useful for fans as well. We built a Shiny application to showcase our models for the general public. Now any fan can put in the situation of the game and guess the next pitch. The exciting part is getting to follow the game in real time with pitch predictions. After each pitch, users can update the situation they have set in the app to see the new prediction of the upcoming pitch. The additional immersion in the game and seeing if you can guess the next pitch correctly can add to the baseball experience. For baseball enthusiasts, a major attraction to baseball is the "game within the game", which is figuring out how to best respond to each state of the game. And a big part of that is the batter trying to guess what pitch is coming and the pitcher trying to throw a pitch that the batter might not expect. Using our models, the enthusiast can become part of the mental chess that goes on in between pitches. For the casual, for which baseball has lost popularity in recent years, the app can be used to break up the lull that occurs between pitches. Therefore, we believe our models have both practical merit and entertainment value for the baseball fan.  

## Appendix A

```{r}
types_Trevor_Rosenthal
types_Felix_Hernandez
types_Chris_Archer
types_Zach_Britton
types_Wade_Davis
types_Dallas_Keuchel
types_Corey_Kluber
types_Luke_Gregerson
types_David_Price
types_Max_Scherzer
types_Aroldis_Chapman
types_Clayton_Kershaw
types_Madison_Bumgarner
types_Sonny_Gray
types_Huston_Street
types_Brad_Boxberger
types_Zack_Greinke
types_Shawn_Tolleson
types_Jordan_Zimmermann
types_Jacob_deGrom
types_Gerrit_Cole
types_Mark_Melancon
types_Jake_Arrieta
types_Andrew_Miller
types_Stephen_Strasburg
types_Collin_McHugh
types_Michael_Wacha
types_Chris_Sale
types_Zack_Greinke
types_Yu_Darvish
```

## Appendix B

```{r tables, include=TRUE}
overall_performance_all
```

## Appendix C

```{r}
by_pitch_performance_Trevor_Rosenthal
by_pitch_performance_Felix_Hernandez
by_pitch_performance_Chris_Archer
by_pitch_performance_Zach_Britton
by_pitch_performance_Wade_Davis
by_pitch_performance_Dallas_Keuchel
by_pitch_performance_Corey_Kluber
by_pitch_performance_Luke_Gregerson
by_pitch_performance_David_Price
by_pitch_performance_Max_Scherzer
by_pitch_performance_Aroldis_Chapman
by_pitch_performance_Clayton_Kershaw
by_pitch_performance_Madison_Bumgarner
by_pitch_performance_Sonny_Gray
by_pitch_performance_Huston_Street
by_pitch_performance_Brad_Boxberger
by_pitch_performance_Zack_Greinke
by_pitch_performance_Shawn_Tolleson
by_pitch_performance_Jordan_Zimmermann
by_pitch_performance_Jacob_deGrom
by_pitch_performance_Gerrit_Cole
by_pitch_performance_Mark_Melancon
by_pitch_performance_Jake_Arrieta
by_pitch_performance_Andrew_Miller
by_pitch_performance_Stephen_Strasburg
by_pitch_performance_Collin_McHugh
by_pitch_performance_Michael_Wacha
by_pitch_performance_Chris_Sale
by_pitch_performance_Zack_Greinke
by_pitch_performance_Yu_Darvish
```

## Appendix D

Datasets: Due to the size of datasets in this folder, we gitignore all .csv files. You can find some of the data in archive.zip. However, download pitches.zip from one of the below links to get pitches data:

- data is from here: http://inalitic.com/datasets/mlb%20pitch%20data.html
- pitchers data dropbox: https://www.dropbox.com/s/9gyz3ujwx7jsh5j/pitches.zip?dl=0
- data is built from here: https://www.kaggle.com/datasets/pschale/mlb-pitch-data-20152018
- detailed info on the data: https://docs.google.com/document/d/1ztD20pt5K0HUi2EcJHT4SYdOZw9YPYhtLUmi8BpInuA/edit?pref=2&pli=1#heading=h.mnao9thv84r1
- hitting here: https://www.baseball-reference.com/leagues/majors/2015-standard-batting.shtml?sr&utm_source=direct&utm_medium=Share&utm_campaign=ShareTool#players_standard_batting

## Appendix E

Glossary of Pitch Types:

CH = Changeup
CU = Curveball
EP = Eephus
FA = Fastball
FC = Cutter
FF = Four seam Fastball
FS = Splitter
FT = Two seam Fastball
FO = Forkball
IN = Intent ball
KC = Knuckle ball Curve
KN = Knuckle ball
PO = Pitch Out
SC = Screwball
SI = Sinker
SL = Slider

