---
title: "Predi-Pitch: Snack on This Curve Whenever You Want"
author: "David Scolari, Harrison Snell, Brandon Williams"
date: "4/13/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(here)
path <- here()

source(file.path(path, 'code', 'dugout.R'))

load(file = file.path(path, 'output', 'tables', "overall_performance_all.RDs"))


foreach(id = 1:30) %do% {
  # make filenames according to loop index
  fname_type <- paste0("types", id, ".RDs")
  fname_zones <- paste0("zones", id, ".RDs")
  fname_bypitch <- paste0("by_pitch_performance", id, ".RDs")

  # load table outputs according to loop index
  load(file = file.path(path, 'output', 'tables', fname_type))
  load(file = file.path(path, 'output', 'tables', fname_zones))
  load(file = file.path(path, 'output', 'tables', fname_bypitch))
  }
```

## Introduction
Gentlemen, pitch predicting is a very important part of baseball. So important that teams will do almost anything for insight on what the opposing team's pitcher is going to throw. Other than illicit methods, hitters are trained to use information on a pitcher's previous pitch throwing behvior to guess which pitch is coming next. We can model this deciscion making process with machine learning techniques. 

By modeling pitch prediction, we can learn what information is relevant to a pitcher's pitch selection. We can also better understand what makes a pitcher more unpredictable than other. It's gonna be good

In the following report will talk about why pithes are different, share evidence that pitchers have different strategies from at bat to at bat, and describe the engineering of our model, which is basically a trashcan .

## Pitches are different you dumb dumb idiot
Harrison's graphs

## Pichers be sequencing
AB networks

## Our Random Forrest 
describe the features, describe the dream 

## what it all means, probably nothing 

## Appendix
```{r tables, include=TRUE}
overall_performance_all

types_Trevor_Rosenthal
types_Felix_Hernandez
types_Chris_Archer
types_Zach_Britton
types_Wade_Davis
types_Dallas_Keuchel
types_Corey_Kluber
types_Luke_Gregerson
types_David_Price
types_Max_Scherzer
types_Aroldis_Chapman
types_Clayton_Kershaw
types_Madison_Bumgarner
types_Sonny_Gray
types_Huston_Street
types_Brad_Boxberger
types_Zack_Greinke
types_Shawn_Tolleson
types_Jordan_Zimmermann
types_Jacob_deGrom
types_Gerrit_Cole
types_Mark_Melancon
types_Jake_Arrieta
types_Andrew_Miller
types_Stephen_Strasburg
types_Collin_McHugh
types_Michael_Wacha
types_Chris_Sale
types_Zack_Greinke
types_Yu_Darvish

by_pitch_performance_Trevor_Rosenthal
by_pitch_performance_Felix_Hernandez
by_pitch_performance_Chris_Archer
by_pitch_performance_Zach_Britton
by_pitch_performance_Wade_Davis
by_pitch_performance_Dallas_Keuchel
by_pitch_performance_Corey_Kluber
by_pitch_performance_Luke_Gregerson
by_pitch_performance_David_Price
by_pitch_performance_Max_Scherzer
by_pitch_performance_Aroldis_Chapman
by_pitch_performance_Clayton_Kershaw
by_pitch_performance_Madison_Bumgarner
by_pitch_performance_Sonny_Gray
by_pitch_performance_Huston_Street
by_pitch_performance_Brad_Boxberger
by_pitch_performance_Zack_Greinke
by_pitch_performance_Shawn_Tolleson
by_pitch_performance_Jordan_Zimmermann
by_pitch_performance_Jacob_deGrom
by_pitch_performance_Gerrit_Cole
by_pitch_performance_Mark_Melancon
by_pitch_performance_Jake_Arrieta
by_pitch_performance_Andrew_Miller
by_pitch_performance_Stephen_Strasburg
by_pitch_performance_Collin_McHugh
by_pitch_performance_Michael_Wacha
by_pitch_performance_Chris_Sale
by_pitch_performance_Zack_Greinke
by_pitch_performance_Yu_Darvish

```